---
title: 'Text analysis of accepted full papers at GIScience conferences 2012, 2014, 2016, and 2018'
author: "Daniel Nüst, Opening Reproducible Research (o2r), Institute for Geoinformatics, University of Münster"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    toc: yes
params:
  with_sp: FALSE
---

## Introduction

This document is an exploratory analysis of all accepted full papers, and posters at the [GIScience conference series](https://www.giscience.org/).

- 2018: Tenth International Conference on Geographic Information Science, Melbourne, Australia
- 2016: Ninth International Conference on Geographic Information Science, Sep 27, 2016 - Sep 30, 2016, 	Montreal, Canada
- 2014: Eighth International Conference on Geographic Information Science, Sep 24, 2014 - Sep 26, 2014, 	Vienna, Austria
- 2012: Seventh International Conference on Geographic Information Science, Sep 18, 2012 - Sep 21, 2012, Columbus, Ohio, USA

The analysis is based on the text analysis published in _"Reproducible research and GIScience: an evaluation using AGILE conference papers"_ ([https://doi.org/10.7717/peerj.5072](https://doi.org/10.7717/peerj.5072)).

```{r load_libraries, message=FALSE, warning=FALSE, include=FALSE}
library("here")
library("pdftools")
library("stringr")
library("tidyverse")
library("tidytext")
library("wordcloud")
library("RColorBrewer")
library("grid")
library("gridBase")
library("gridExtra")
library("kableExtra")
library("quanteda")

# for deterministic cloud rendering
set.seed(nchar("International Conference on Geographic Information Science"))
```

## Load data

**List of proceedings**

- Proceedings 10th International Conference on Geographic Information Science (GIScience 2018). 2018. In Winter, S., Griffin, A., Sester, M. (Eds.), LIPICS Vol. 114. ISBN 978-3-95977-083-5. http://www.dagstuhl.de/dagpub/978-3-95977-083-5
- Geographic Information Science. 2016. In J. A. Miller, D. O’Sullivan, & N. Wiegand (Eds.), Lecture Notes in Computer Science. Springer International Publishing. https://doi.org/10.1007/978-3-319-45738-3
- Geographic Information Science. 2014. In M. Duckham, E. Pebesma, K. Stewart, & A. U. Frank (Eds.), Lecture Notes in Computer Science. Springer International Publishing. https://doi.org/10.1007/978-3-319-11593-1
- Geographic Information Science. 2012. In N. Xiao, M.-P. Kwan, M. F. Goodchild, & S. Shekhar (Eds.), Lecture Notes in Computer Science. Springer Berlin Heidelberg. https://doi.org/10.1007/978-3-642-33024-7

See the full list of proceedings for 2016 and older at the publisher website: [https://link.springer.com/conference/giscience](https://link.springer.com/conference/giscience).

**Note:** The 2018 proceedings include the short papers in the same document.
For comparability, only the full papers are taken into account for the analysis below.

```{r input_files, echo=FALSE, cache=TRUE}
cat("Loading proceedigns from ", here::here())
data_path <- here::here("proceedings")
proceedings <- c(
  #"2002" = "geographic-information-science-2002.pdf",
  #"2004" = "geographic-information-science-2004.pdf",
  #"2006" = "geographic-information-science-2006.pdf",
  #"2008" = "geographic-information-science-2008.pdf",
  #"2010" = "geographic-information-science-2010.pdf",
  "2012" = "10.1007_978-3-642-33024-7.pdf",
  "2014" = "10.1007_978-3-319-11593-1.pdf",
  "2016" = "10.1007_978-3-319-45738-3.pdf",
  "2018" = "lipics-vol114-giscience2018-complete.pdf"
)
proceedings_files <- file.path(data_path, proceedings)
names(proceedings_files) <- names(proceedings)
```

Add the PDFs to a directory called ` `r data_path` ` next to the file `giscience-historic-text-analysis.Rmd` (this file).
The proceedings of the papers are not openly available for the years 2012 to 2016.
You can contact the original paper authors and ask for the test dataset to reproduce the full analysis.
Alternatively, you can download the 2018 proceedings from the LIPIcs website (Open Access; [direct PDF link](https://drops.dagstuhl.de/opus/volltexte/lipics-complete/lipics-vol114-giscience2018-complete.pdf)) and conduct the analysis with that subset of the data.
For the analysis below the following input files were used:

```{r list_files, echo=FALSE}
knitr::kable(tibble(year = names(proceedings), file = proceedings)) %>%
  kableExtra::kable_styling("striped", full_width = FALSE)
```

```{r data_download_drive, eval=FALSE, echo=FALSE}
dir.create(data_path, showWarnings = FALSE)

library("googledrive")
drive_dir <- googledrive::drive_get("https://drive.google.com/drive/folders/17EUtM_zCx1gQMea1MHN_5XSVrssxv9GA")
drive_dir_contents <- googledrive::drive_ls(drive_dir)
for (i in rownames(drive_dir_contents)) {
  current <- drive_dir_contents[i,]
  if(endsWith(current$name, ".pdf"))
    googledrive::drive_download(as_id(current$id), file.path(data_path, current$name))
}
```

The text is extracted from PDFs and it is processed to create a [tidy](https://www.jstatsoft.org/article/view/v059i10) data structure without [stop words](https://en.wikipedia.org/wiki/Stop_words).
The stop words include specific words, such as `university`, which is included in many pages, abbreviations such as `e.g.`, and terms particular to scientific articles, such as `figure`.
Also all numeric literas are removed from the word list.

```{r load_files, echo=FALSE, cache=TRUE}
texts <- lapply(proceedings_files, pdftools::pdf_text)

if(params$with_sp) {
  texts[["2018-sp"]] <- texts[["2018"]][c(283:length(texts[["2018"]]))]
  proceedings_files <- c(proceedings_files, `2018-sp` = proceedings_files[[4]])
}

texts[["2018"]] <- texts[["2018"]][c(1:282)]

texts <- unlist(lapply(texts, stringr::str_c, collapse = TRUE))

tidy_texts <- tibble::tibble(year = names(texts),
                             path = proceedings_files,
                             text = texts)

# create a table of all words
all_words <- tidy_texts %>%
  dplyr::select(year, text) %>%
  tidytext::unnest_tokens(word, text)

# remove stop words and remove numbers
my_stop_words <- tibble::tibble(
  word = c(
    "et",
    "al",
    "fig",
    "e.g",
    "i.e",
    "http",
    "ing",
    "pp",
    "figure",
    "based",
    "conference",
    "university",
    "table"
  ),
  lexicon = "giscience"
)

all_stop_words <- stop_words %>%
  dplyr::bind_rows(my_stop_words)
suppressWarnings({
  no_numbers <- all_words %>%
    dplyr::filter(is.na(as.numeric(word)))
})

no_stop_words <- no_numbers %>%
  dplyr::anti_join(all_stop_words, by = "word")

total_words = nrow(no_numbers)
after_cleanup = nrow(no_stop_words)
```

About `r round(after_cleanup/total_words * 100)`&nbsp;% of the words are considered stop words.
The following tables shows how many non-stop words each conference year has, sorted by number of non-stop words.

```{r stop_words, echo=FALSE, message=FALSE, warning=FALSE}
nonstopwords_per_year <- no_stop_words %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(words = n()) %>%
  dplyr::arrange(desc(words)) %>%
  dplyr::rename(`non-stop words` = words)

words_per_year <- no_numbers %>%
  dplyr::group_by(year) %>%
  dplyr::summarise(words = n()) %>%
  dplyr::arrange(desc(words)) %>%
  dplyr::rename(`all words` = words)

dplyr::inner_join(nonstopwords_per_year, words_per_year, by = "year") %>%
  dplyr::bind_rows(tibble(year = "Total",
                   `non-stop words` = sum(nonstopwords_per_year$`non-stop words`),
                   `all words` = sum(words_per_year$`all words`))) %>%
  knitr::kable() %>%
  kableExtra::kable_styling("striped", full_width = FALSE) %>%
  kableExtra::row_spec(nrow(nonstopwords_per_year) + 1, bold = TRUE)
```

## Top wordstems and word clouds

```{r params, include=FALSE}
# chosen manually
minimum_occurence <- 100
max_words <- 100
```

The following table shows the number of occurence for the `r max_words` most occuring wordstems across all proceedings.

```{r cloud_and_top_words, echo=FALSE}
top_words <- no_stop_words %>%
  dplyr::group_by(word) %>%
  dplyr::tally() %>%
  dplyr::arrange(desc(n)) %>%
  head(20)

wordstems <- no_stop_words %>%
  dplyr::mutate(wordstem = quanteda::char_wordstem(no_stop_words$word))

countYearsUsingWordstem <- function(the_word) {
  sapply(the_word, function(w) {
    wordstems %>%
      dplyr::filter(wordstem == w) %>%
      dplyr::group_by(year) %>%
      dplyr::count() %>%
      nrow
  })
}

top_wordstems <- wordstems %>%
  dplyr::group_by(wordstem) %>%
  dplyr::tally() %>%
  dplyr::arrange(desc(n)) %>%
  head(n = max_words) %>%
  dplyr::mutate(`year w/ wordstem` = countYearsUsingWordstem(wordstem)) %>%
  tibble::add_column(place = c(1:nrow(.)), .before = 0)

cloud_wordstems <- wordstems %>%
  dplyr::group_by(year, wordstem) %>%
  dplyr::tally() %>%
  dplyr::arrange(desc(n))

top_wordstems %>%
  head(n = max_words) %>%
  knitr::kable() %>%
  kableExtra::kable_styling("striped", full_width = FALSE) %>%
  kableExtra::scroll_box(height = "300px")
```

The following clouds and table are based on word stems extracted with a stemming algorithm from package [`quanteda`](https://cran.r-project.org/package=quanteda).
Words must occur at least `r minimum_occurence` times to be included in the cloud.
Each cloud has a maximum of `r max_words` words.

```{r wordclouds, echo=FALSE, dpi=300, fig.width=12, fig.height=10, fig.cap="World clouds of full papers per conference year (clockwise starting top left: 2012, 2014, 2016, 2018)"}
par(mfrow=c(3,3))

for (the_year in names(proceedings)) {
  cloud_words <- cloud_wordstems %>%
    dplyr::filter(year == the_year) %>%
    dplyr::filter(n >= minimum_occurence) %>%
    head(n = max_words)
  wordcloud(cloud_words$wordstem, cloud_words$n,
            random.order = FALSE,
            fixed.asp = FALSE,
            rot.per = 0,
            color = brewer.pal(8,"Dark2"))
}
```

## Reproducible research-related keywords in GIScience papers

The following tables lists how often terms related to reproducible research appear in each document.
The detection matches full words using regex option `\b`.

- reproduc (`reproduc.*`, reproducibility, reproducible, reproduce, reproduction)
- replic (`replicat.*`, i.e. replication, replicate)
- repeatab (`repeatab.*`, i.e. repeatability, repeatable)
- software
- (pseudo) code/script(s) [column name _code_]
- algorithm (`algorithm.*`, i.e. algorithms, algorithmic)
- process (`process.*`, i.e. processing, processes, preprocessing)
- data (`data.*`, i.e. dataset(s), database(s))
- result(s) (`results?`)
- repository(ies) (`repositor(y|ies)`)
- collaboration platforms (`git(hub|lab)`)

```{r keywords_per_year, echo=FALSE, warning=FALSE}
tidy_texts_lower <- stringr::str_to_lower(tidy_texts$text)
word_counts <- tibble::tibble(
  year = tidy_texts$year,
  `words` = str_count(tidy_texts_lower, "\\b.*\\b"),
  `reproduc..` = str_count(tidy_texts_lower, "\\breproduc.*\\b"),
  `replic..` = str_count(tidy_texts_lower, "\\breplicat.*\\b"),
  `repeatab..` = str_count(tidy_texts_lower, "\\brepeatab.*\\b"),
  `code` = str_count(tidy_texts_lower,
    "(\\bcode\\b|\\bscript.*\\b|\\bpseudo\ code\\b)"),
  software = str_count(tidy_texts_lower, "\\bsoftware\\b"),
  `algorithm(s)` = str_count(tidy_texts_lower, "\\balgorithm.*\\b"),
  `(pre)process..` = str_count(tidy_texts_lower, 
                "(\\bprocess.*\\b|\\bpreprocess.*\\b|\\bpre-process.*\\b)"),
  `data.*` = str_count(tidy_texts_lower, "\\bdata.*\\b"),
  `result(s)` = str_count(tidy_texts_lower, "\\bresults?\\b"),
  `repository/ies` = str_count(tidy_texts_lower, "\\brepositor(y|ies)\\b"),
  #`repos` = str_count(tidy_texts_lower, "\\bzenodo|figshare|osf|dryad\\b"),
  `github/lab` = str_count(tidy_texts_lower, "\\bgit(hub|lab)\\b")
)

word_counts_sums <- rbind(word_counts,
                          word_counts %>% 
                            dplyr::summarise_if(is.numeric, funs(sum)) %>%
                            tibble::add_column(year = "Total", .before = 0))

word_counts_sums %>%
  knitr::kable() %>%
  kableExtra::kable_styling("striped", font_size = 10, bootstrap_options = "condensed")  %>%
  kableExtra::row_spec(0, font_size = "x-small", bold = T)  %>%
  kableExtra::row_spec(nrow(word_counts_sums), bold = T)
```

## License & Metadata

This document is licensed under a [Creative Commons Attribution 4.0 International License](https://creativecommons.org/licenses/by/4.0/).
All contained code is licensed under the [Apache License 2.0](https://choosealicense.com/licenses/apache-2.0/).
This document is versioned in a public [git](https://git-scm.com/) repository, [https://github.com/nuest/reproducible-research-at-giscience](https://github.com/nuest/reproducible-research-at-giscience).

**Runtime environment description:**

```{r session_info, echo=FALSE}
devtools::session_info(include_base = TRUE)
```

```{r upload_to_drive, eval=FALSE, include=FALSE}
# upload the HTML and Rmd file to the authoring team's shared folder
library("googledrive")
googledrive::drive_auth()
googledrive::drive_put("giscience-historic-text-analysis.Rmd", path = "https://drive.google.com/drive/folders/17EUtM_zCx1gQMea1MHN_5XSVrssxv9GA/")
googledrive::drive_put("giscience-historic-text-analysis.html", path = "https://drive.google.com/drive/folders/17EUtM_zCx1gQMea1MHN_5XSVrssxv9GA/")
```
