---
title: 'Analysis and visualisations for "Reproducible research and GIScience: an evaluation using GIScience conference papers"'
author: "Daniel Nüst"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    toc: yes
  html_document:
    df_print: paged
    toc: yes
urlcolor: blue
# add lscape package to support kableExtra::landscape() for PDF output
header-includes: |
  \usepackage{lscape}
---

This document is an exploratory analysis of all accepted full papers, and posters at the [GIScience conference series](https://www.giscience.org/).

- 2018: Tenth International Conference on Geographic Information Science, Melbourne, Australia
- 2016: Ninth International Conference on Geographic Information Science, Sep 27, 2016 - Sep 30, 2016, 	Montreal, Canada
- 2014: Eighth International Conference on Geographic Information Science, Sep 24, 2014 - Sep 26, 2014, 	Vienna, Austria
- 2012: Seventh International Conference on Geographic Information Science, Sep 18, 2012 - Sep 21, 2012, Columbus, Ohio, USA

The analysis is based on the text analysis published in _"Reproducible research and GIScience: an evaluation using AGILE conference papers"_ ([https://doi.org/10.7717/peerj.5072](https://doi.org/10.7717/peerj.5072)), see [https://github.com/nuest/reproducible-research-and-giscience](https://github.com/nuest/reproducible-research-and-giscience).

## Reproduce paper

_If you do not have the original data or do not download the data, you cannot reproduce the text analysis part of the paper, i.e. wordcloud and terms frequency analysis._ **You can still reproduce the other figures**.

To create the PDF of the reproducibility package based on this document you can run the following commandsin a new R session after completing the prerequisites with the original paper corpus data.
If you have problems rendering the PDF and execute each chunk independently, _skip the following chunk_.

```{r render_with_rmarkdown,eval=FALSE}
require("knitr")
require("rmarkdown")
rmarkdown::render("giscience-paper-reproducibility.Rmd", output_format = "pdf_document")
```

## Software dependencies

This document does not install the required R packages by default.
You can run the script `install.R` to install all required dependencies on a new R installation, or use `install.packages(..)` to install missing R packages.

```{r install_r, eval=FALSE}
source("install.R")
```

## Dependencies and code

The **text analysis** is based the R package [`tidytext`](https://cran.r-project.org/package=tidytext) from the [`tidyverse`](https://www.tidyverse.org/) suite of packages and uses the [`dplyr`](http://dplyr.tidyverse.org/) grammar.
Read the [`tidytext` tutorial](https://cran.r-project.org/web/packages/tidytext/vignettes/tidytext.html) to learn about the used functions and concepts.
The **plots and tables** of survey data and evaluation use the packages [`ggplot2`](http://ggplot2.tidyverse.org/), [`knitr::kable()`](https://yihui.name/knitr/), [`huxtable`](https://hughjonesd.github.io/huxtable/), and [`kableExtra`](https://cran.r-project.org/package=kableExtra).
Required libraries and runtime environment description are as follows.

```{r load_libraries, echo=TRUE, message=FALSE, warning=FALSE}
library("pdftools")
library("stringr")
library("tidyverse")
library("knitr")
library("tidytext")
library("wordcloud")
library("RColorBrewer")
library("readr")
library("ggplot2")
library("rvest")
library("jsonlite")
library("reshape2")
library("ggthemes")
library("grid")
library("gridBase")
library("gridExtra")
library("kableExtra")
library("devtools")
library("rlang")
library("huxtable")
library("here")
library("httr")
library("googledrive")
library("googlesheets")
```

```{r set_seed}
# for deterministic rendering of the wordcloud
set.seed(nchar("International Conference on Geographic Information Science"))
```

## Load data

```{r paths, echo=FALSE}
data_path <- here::here("papers")
```

The data for the analysis is required in form of a directory with PDF files.
Add the PDFs to a directory called ` `r data_path` ` (path automatically inserted here based on above variable) next to the file `giscience-paper-reproducibility.Rmd` (this file).
You can contact the original paper authors and ask for the test dataset to reproduce the full analysis.
Alternatively, you can download the 2018 papers from the Open Access proceedings at [https://drops.dagstuhl.de/opus/portals/lipics/index.php?semnr=16081](https://drops.dagstuhl.de/opus/portals/lipics/index.php?semnr=16081) and conduct the analysis with that subset of the data.

```{r data_download_drive, eval=FALSE, echo=FALSE}
dir.create(data_path)

drive_dir <- drive_get("https://drive.google.com/drive/folders/17EUtM_zCx1gQMea1MHN_5XSVrssxv9GA")
drive_dir_contents <- drive_ls(drive_dir)
for (i in rownames(drive_dir_contents)) {
  current <- drive_dir_contents[i,]
  if(endsWith(current$name, ".pdf"))
    drive_download(as_id(current$id), file.path(data_path, current$name))
}
```

The text is extracted from PDFs and it is processed to create a [tidy](https://www.jstatsoft.org/article/view/v059i10) data structure without [stop words](https://en.wikipedia.org/wiki/Stop_words).
The stop words include specific words, such as `university`, which is included in many pages, abbreviations such as `e.g.`, and terms particular to scientific articles, such as `figure`.
Also all numeric literals are removed from the word list.

```{r load_files, echo=FALSE, cache=TRUE}
proceedings <- c("2012" = "10.1007_978-3-642-33024-7.pdf",
  "2014" = "10.1007_978-3-319-11593-1.pdf",
  "2016" = "10.1007_978-3-319-45738-3.pdf",
  "2018" = "lipics-vol114-giscience2018-complete.pdf")
proceedings_files <- file.path(data_path, proceedings)

texts <- lapply(proceedings_files, pdf_text)
texts <- unlist(lapply(texts, str_c, collapse = TRUE))
infos <- lapply(proceedings_files, pdf_info)

tidy_texts <- tibble(year = names(proceedings),
                     path = proceedings_files,
                     text = texts,
                     pages = map_chr(infos, function(info) {info$pages}))

# create a table of all words
all_words <- tidy_texts %>%
  select(year,
         text) %>%
  unnest_tokens(word, text)

# remove stop words and remove numbers
my_stop_words <- tibble(
  word = c(
    "et",
    "al",
    "fig",
    "e.g",
    "i.e",
    "http",
    "ing",
    "pp",
    "figure",
    "based",
    "conference",
    "university",
    "table"
  ),
  lexicon = "agile"
)

all_stop_words <- stop_words %>%
  bind_rows(my_stop_words)
suppressWarnings({
  no_numbers <- all_words %>%
    filter(is.na(as.numeric(word)))
})

no_stop_words <- no_numbers %>%
  anti_join(all_stop_words, by = "word")

total_words = nrow(no_numbers)
after_cleanup = nrow(no_stop_words)
```

About `r round(after_cleanup/total_words * 100)`&nbsp;% of the words are considered stop words.
The following tables shows how many non-stop words each conference year has, sorted by number of non-stop words.

```{r stop_words, echo=FALSE, message=FALSE, warning=FALSE}
nonstopwords_per_doc <- no_stop_words %>%
  group_by(year) %>%
  summarise(words = n()) %>%
  arrange(desc(words)) %>%
  rename(`non-stop words` = words)

words_per_doc <- no_numbers %>%
  group_by(year) %>%
  summarise(words = n()) %>%
  arrange(desc(words)) %>%
  rename(`all words` = words)

inner_join(nonstopwords_per_doc, words_per_doc) %>%
  bind_rows(tibble(year = "Total",
                   `non-stop words` = sum(nonstopwords_per_doc$`non-stop words`),
                   `all words` = sum(words_per_doc$`all words`))) %>%
  kable() %>%
  kable_styling("striped", full_width = FALSE) %>%
  row_spec(nrow(nonstopwords_per_doc) + 1, bold = TRUE)
```

## Text analysis and visualisation

### Top wordstems and word clouds

```{r params, include=FALSE}
# chosen manually
minimum_occurence <- 60
max_words <- 120
```

The following clouds and table are based on word stems extracted with a stemming algorithm from package [`quanteda`](https://cran.r-project.org/package=quanteda).
Words must occur at least `r minimum_occurence` times to be included.
Each cloud has a maximum of `r max_words` words.

```{r cloud_and_top_words, echo=FALSE}
top_words <- no_stop_words %>%
  group_by(word) %>%
  tally %>%
  arrange(desc(n)) %>%
  head(20)

wordstems <- no_stop_words %>%
  mutate(wordstem = quanteda::char_wordstem(no_stop_words$word))

countYearsUsingWordstem <- function(the_word) {
  sapply(the_word, function(w) {
    wordstems %>%
      filter(wordstem == w) %>%
      group_by(year) %>%
      count %>%
      nrow
  })
}

top_wordstems <- wordstems %>%
  group_by(wordstem) %>%
  tally %>%
  arrange(desc(n)) %>%
  head(n = 100) %>%
  mutate(`year w/ wordstem` = countYearsUsingWordstem(wordstem)) %>%
  add_column(place = c(1:nrow(.)), .before = 0)

cloud_wordstems <- wordstems %>%
  group_by(year, wordstem) %>%
  tally %>%
  arrange(desc(n))

top_wordstems %>%
  head(n = 100) %>%
  kable() %>%
  kable_styling("striped", full_width = FALSE) %>%
  scroll_box(height = "240px")
```

```{r wordclouds, echo=FALSE, dpi=300, fig.width=12, fig.height=10, fig.cap="World clouds per conference year (clockwise starting top left: 2012, 2014, 2016, 2018)"}
par(mfrow=c(2,2))

for (the_year in names(proceedings)) {
  cloud_words <- cloud_wordstems %>%
    filter(year == the_year) %>%
    filter(n >= minimum_occurence) %>%
    head(n = max_words)
  wordcloud(cloud_words$wordstem, cloud_words$n,
            random.order = FALSE,
            fixed.asp = FALSE,
            rot.per = 0,
            color = brewer.pal(8,"Dark2"))
}
```

### Reproducible research-related keywords in GIScience papers

The following tables lists how often terms related to reproducible research appear in each document.
The detection matches full words using regex option `\b`.

- reproduc (`reproduc.*`, reproducibility, reproducible, reproduce, reproduction)
- replic (`replicat.*`, i.e. replication, replicate)
- repeatab (`repeatab.*`, i.e. repeatability, repeatable)
- software
- (pseudo) code/script(s) [column name _code_]
- algorithm (`algorithm.*`, i.e. algorithms, algorithmic)
- process (`process.*`, i.e. processing, processes, preprocessing)
- data (`data.*`, i.e. dataset(s), database(s))
- result(s) (`results?`)
- repository(ies) (`repositor(y|ies)`)
- collaboration platforms (`git(hub|lab)`)

```{r keywords_per_paper, echo=FALSE, warning=FALSE}
tidy_texts_lower <- str_to_lower(tidy_texts$text)
word_counts <- tibble(
  year = tidy_texts$year,
  `reproduc..` = str_count(tidy_texts_lower, "\\breproduc.*\\b"),
  `replic..` = str_count(tidy_texts_lower, "\\breplicat.*\\b"),
  `repeatab..` = str_count(tidy_texts_lower, "\\brepeatab.*\\b"),
  `code` = str_count(tidy_texts_lower,
    "(\\bcode\\b|\\bscript.*\\b|\\bpseudo\ code\\b)"),
  software = str_count(tidy_texts_lower, "\\bsoftware\\b"),
  `algorithm(s)` = str_count(tidy_texts_lower, "\\balgorithm.*\\b"),
  `(pre)process..` = str_count(tidy_texts_lower, 
                "(\\bprocess.*\\b|\\bpreprocess.*\\b|\\bpre-process.*\\b)"),
  `data.*` = str_count(tidy_texts_lower, "\\bdata.*\\b"),
  `result(s)` = str_count(tidy_texts_lower, "\\bresults?\\b"),
  `repository/ies` = str_count(tidy_texts_lower, "\\brepositor(y|ies)\\b"),
  #`repos` = str_count(tidy_texts_lower, "\\bzenodo|figshare|osf|dryad\\b"),
  `github/lab` = str_count(tidy_texts_lower, "\\bgit(hub|lab)\\b")
)

word_counts_sums <- rbind(word_counts,
                          word_counts %>% 
                            summarise_if(is.numeric, funs(sum)) %>%
                            add_column(year = "Total", .before = 0))

word_counts_sums %>%
  kable() %>%
  kable_styling("striped", font_size = 12, bootstrap_options = "condensed")  %>%
  row_spec(0, font_size = "x-small", bold = T)  %>%
  row_spec(nrow(word_counts_sums), bold = T)
```

## Reproduciblity assessment

```{r evaldata_file}
assessment_file <- "paper_assessment.csv"
```

```{r evaldata_download, echo=FALSE}
googlesheets::gs_download(from = "https://docs.google.com/spreadsheets/d/1Df0zX5yqDuDSCzHHVaucNtThseHL5L32U2sWTPhyl0I",
                          to = here::here(assessment_file))
```

The following plots are based on the file `r assessment_file`, the result from the manual reproducibility assessment.

```{r load_evaldata,warning=FALSE}
category_levels <- c("0", "1", "2", "3")
paper_evaluation_raw <- read_csv(assessment_file, 
    col_types = cols(
      paper = col_skip(),
      title = col_skip(),
      `Notes Reviewer` = col_skip(),
      `computational environment` = col_factor(levels = category_levels),
      `input data` = col_factor(levels = category_levels),
      `method/analysis/processing` = col_factor(levels = category_levels),
      preprocessing = col_factor(levels = category_levels),
      results = col_factor(levels = category_levels),
      X12 = col_skip(),
      X14 = col_skip(),
      `Notes Reviewer` = col_skip(),
      `Author comment` = col_skip()
      ),
    na = "NA")
categoryColumns <- c("input data", 
                     "preprocessing",
                     "method/analysis/processing",
                     "computational environment",
                     "results")
```

```{r corpus_table_with_small_font_for_latex}
options(knitr.kable.NA = '-')
kable(paper_evaluation_raw %>% 
        select(-matches("reviewer")) %>%
        mutate(`short paper` = if_else(`short paper` == TRUE, "X", "")),
      format = "latex", # change output format to "html" when running the chunk manually
      #format = "html",
      booktabs = TRUE,
      caption = paste0("Reproducibility levels for paper corpus; ",
                       "'-' is category not available")) %>%
  kable_styling(latex_options = "scale_down")
```

\newpage

## Conceptual papers

```{r conceptual_papers,warning=FALSE}
paper_evaluation <- paper_evaluation_raw %>%
  # add year column
  mutate(year = as.numeric(str_extract(author, "[0-9]+"))) %>%
  # create new attribute for conceptual papers
  mutate(conceptual = is.na(`input data`) 
         & is.na(preprocessing) 
         & is.na(`method/analysis/processing`) 
         & is.na(`computational environment`) 
         & is.na(results))

count_conceptual <- nrow(paper_evaluation %>% 
                           filter(conceptual))
count_mixed <- nrow(paper_evaluation %>% 
                      filter(is.na(`input data`) 
                             | is.na(preprocessing) 
                             | is.na(`method/analysis/processing`) 
                             | is.na(`computational environment`) 
                             | is.na(results)))
```

`r count_conceptual` papers are purely conceptual (all categories have value `NA`).
These are not included in the following statistics.

`r count_mixed` papers are partically conceptual (at least one category has a value of `NA`).
These are evaluated.

`r paper_evaluation %>% filter(is.na(preprocessing)) %>% count() %>% .$n` papers are not applicable for preprocessing criterion.

## Table: Statistics of reproducibility levels per criterion

```{r summary_evaldata}
evaldata_numeric <- paper_evaluation %>%
  # must convert factors to numbers to calculate the mean and median
  mutate_if(is.factor, funs(as.integer(as.character(.))))

summary(evaldata_numeric[,categoryColumns])

# apply summary independently to format as table
summaries <- sapply(evaldata_numeric[,categoryColumns], summary)
exclude_values_summary <- c("1st Qu.", "3rd Qu.")
kable(subset(summaries, !(rownames(summaries) %in% exclude_values_summary)), 
      digits = 2,
      col.names = c("input data", "preproc.", "method/analysis/proc.",
                    "comp. env.", "results"),
      caption = paste0("\\label{tab:levels_statistics}Statistics of ",
                       "reproducibility levels per criterion"))
```

The preprocessing has `r sum(!is.na(evaldata_numeric$preprocessing))` values, with `0` and `1` around the "middle" resulting in a fraction as the median.

\newpage

## Figure: Results of reproducibility assessment

```{r Fig3,fig.width=10}
# match the colours to time series plot below
colours <- RColorBrewer::brewer.pal(length(categoryColumns), "Set1")
level_names <- c("0", "1", "2", "3", "NA")

criteriaBarplot = function(data, main, colour) {
  barplot(table(data, useNA = "always"), 
          main = main,
          xlab = "Level", 
          ylim = c(0,25),
          names.arg = level_names,col = colours[colour])
}

par(mfrow = c(1,length(categoryColumns)))
criteriaBarplot(paper_evaluation$`input data`,
                main = "A: Input data", colour = 1)
criteriaBarplot(paper_evaluation$`preprocessing`, 
                main = "B: Preprocessing", colour = 2)
criteriaBarplot(paper_evaluation$`method/analysis/processing`,
                main = "C: Methods/Analysis/\nProcessing", colour = 3)
criteriaBarplot(paper_evaluation$`computational environment`,
                main = "D: Computational\nEnvironment", colour = 4)
criteriaBarplot(paper_evaluation$results,
                main = "E: Results", colour = 5)
```

```{r criteria_numbers}
data_level_zero <- paper_evaluation %>% 
  filter(`input data` == 0) %>% 
  count() %>% .$n

data_level_two <- paper_evaluation %>% 
  filter(`input data` == 2) %>% 
  count() %>% .$n

preprocessing_included <- paper_evaluation %>% 
  filter(!is.na(preprocessing)) %>% 
  count() %>% .$n

methods_and_results_eq_one <- evaldata_numeric %>% 
  filter(`method/analysis/processing` == 1 & results == 1) %>% 
  count() %>% .$n
```

`r data_level_zero` papers have level `0` and `r data_level_two` have level `2` in the data criterion.

`r preprocessing_included` papers include some kind of preprocessing.

`r methods_and_results_eq_one` papers have level `1` in both methods and results criterion.

\newpage

## Table: Mean levels per criterion for full and short papers

```{r summary_evaldata_grouped}
summaries_short_paper <- sapply(evaldata_numeric %>%
                                  filter(`short paper` == TRUE) %>%
                                  select(categoryColumns), summary)
means_short_paper <- subset(summaries_short_paper, rownames(summaries) %in% c("Mean"))
rownames(means_short_paper) <- c("Short papers")
summaries_full_paper <- sapply(evaldata_numeric %>% filter(`short paper` == FALSE) %>%
                                 select(categoryColumns), summary)
means_full_paper <- subset(summaries_full_paper, rownames(summaries) %in% c("Mean"))
rownames(means_full_paper) <- c("Full papers")
```

\small

```{r summary_evaldata_grouped_smallfont_latex}
kable(rbind(means_full_paper, means_short_paper),
      digits = 2,
      col.names = c("input data", "preproc.", "method/analysis/proc.", "comp. env.", "results"),
      caption = paste0("\\label{tab:mean_full_vs_short}",
                       "Mean levels per criterion for full and short papers"))
```

\normalsize

\newpage

## Extra table: Mean levels averaged across criteria over time

```{r evaldata_summary_by_year_mean}
means_years <- evaldata_numeric %>%
  filter(conceptual == FALSE) %>%
  group_by(year) %>%
  summarise(mean = mean(c(`input data`, 
                          preprocessing, 
                          `method/analysis/processing`, 
                          `computational environment`, 
                          `results`),
                        na.rm = TRUE),
            `paper count` = n())

means_years_table <- means_years %>% 
        mutate(mean = round(mean, 2), 
               `paper count` = as.character(`paper count`)) %>%
        mutate(labels = str_c(year, " (n = ", `paper count`, ")")) %>%
        #column_to_rownames("labels") %>%
        select(mean) %>%
        t()
```

\small

```{r summary_by_year_smallfont_latex}
kable(means_years_table,
      caption = "Summarised mean values over all criteria over time")
```

\normalsize

\newpage

## Figure: Mean reproducibility levels per category over time

```{r Fig4,fig.width=10,dpi=300}
evaldata_years <- evaldata_numeric %>%
  filter(conceptual == FALSE) %>%
  filter(year != 2011) %>%
  group_by(year) %>%
  summarise(input = mean(`input data`, na.rm = TRUE),
         preprocessing = mean(preprocessing, na.rm = TRUE),
            method = mean(`method/analysis/processing`, na.rm = TRUE),
            environment = mean(`computational environment`, na.rm = TRUE),
            results = mean(results, na.rm = TRUE))
paper_count_years <- evaldata_numeric %>%
  filter(conceptual == FALSE) %>%
  filter(year != 2011) %>%
  group_by(year) %>%
  summarise(`paper count` = n())

evaldata_years_long <- melt(evaldata_years, id.vars = c("year"))
ggplot(evaldata_years_long, aes(year, value)) +
  geom_bar(aes(fill = variable), position = "dodge", stat = "identity") +
  ylab("mean value of criterion level") + 
  scale_x_continuous(breaks = evaldata_years$year,
                     labels = paste0(paper_count_years$year, 
                                     " (n=", 
                                     paper_count_years$`paper count`, 
                                     ")")) +
  scale_fill_brewer(palette = "Set1", name = "Category") +
  theme_tufte(base_size = 18) +
  theme(legend.position = c(0.15,0.75), 
        legend.text = element_text(size = 14)) +
  ylim(0, 3) +
  stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean, shape = "-", size = 2) +
  stat_summary(fun.y = mean, geom = "line", linetype = "dotted", mapping = aes(group = 1))
```

## License & Metadata

This document is licensed under a [Creative Commons Attribution 4.0 International License](https://creativecommons.org/licenses/by/4.0/).
All contained code is licensed under the [Apache License 2.0](https://choosealicense.com/licenses/apache-2.0/).

**Runtime environment description:**

```{r session_info, echo=FALSE}
devtools::session_info(include_base = TRUE)
```

This document is versioned in a public [git](https://git-scm.com/) repository, [https://github.com/nuest/reproducible-research-at-giscience](https://github.com/nuest/reproducible-research-at-giscience).


```{r upload_to_drive, eval=FALSE, include=FALSE}
# upload the HTML file to the Reproducibility Committee shared folder
# upload the HTML file and source code to the Reproducibility Committee shared folder
drive_put("giscience-paper-reproducibility.html", path = "https://drive.google.com/drive/folders/17EUtM_zCx1gQMea1MHN_5XSVrssxv9GA")
drive_put("giscience-paper-reproducibility.Rmd", path = "https://drive.google.com/drive/folders/17EUtM_zCx1gQMea1MHN_5XSVrssxv9GA")
```
